plyranges: a grammar for transforming genomics data

### 2018-03-26 paper.Rmd
more words

$$
\textit{GenomicRange} := \{  \{\text{seqname}_i, \left[\text{start}_i, \text{end}_{i}\right] , \text{strand}_i\}  : i = 1\dots n \}.
$$
A _GenomicRanges_ object can have an additional property that provides
annotation for each sequence name over a genome. This contstrains the the 
start and end points for a given sequence to lie inside the bounds
of the genome space. Additional values can be bound to the bare 
_IntegerRanges_ or _GenomicRanges_ to form metadata columns.

@@

We have designed every function to perform a single
action on a _Ranges_ object. The `plyranges` DSL implments the core verbs from the _dplyr_ package
and implements context specific verbs for transforming _Ranges_ objects (described in
table ). Each verb preserves the semantics of _Ranges_ object and via
duck typing works with derivatives of the _Ranges_ class. 
Both of these aspects reduce the cognitive load on a new user since most 
manipulations can be performed with a  vocabulary of several verbs, rather 
than having to memorise function names that are nouns. Workflows can be composed
by chaining verbs together into 'sentences' via the forward pipe operator,`%>%` 
(exported from the R package `magrittr` @R-magrittr), which can be read as the 
word 'then'. Overall, this  allows users to write human readable code because 
workflows describe what the code is doing rather than how its doing it.

>>>>

$$
\textit{GenomicRange} := \{  \{\text{seqname}_i, \left[\text{start}_i, \text{end}_{i}\right] , \text{strand}_i\}  : i = 1\dots n \}.
$$
A _GenomicRanges_ object can have an additional property that provides
annotation for each sequence name over a genome. This contstrains the the 
start and end points for a given sequence to lie inside the bounds
of the genome space. Also note that _GenomicRanges_ have one-based 
coordinates rather than zero-based coordinates.  Additional measurements
or annotations can be bound to the bare _IntegerRanges_ or _GenomicRanges_ to
form metadata columns.

@@

We have designed every function to perform a single
action on a _Ranges_ object. The `plyranges` DSL implments the core verbs from the `dplyr` package
and implements context specific verbs for transforming _Ranges_ objects (described in
table ). Each verb preserves the semantics of _Ranges_ object and via
duck typing works with derivatives of the _Ranges_ class. 
Both of these aspects reduce the cognitive load on a new user since most 
manipulations can be performed with a  vocabulary of several verbs, rather 
than having to memorise function names that are nouns. Moreover, this approach 
strongly constrasts the `GenomicRanges/IRanges` OO interface, which emphasises 
the use of setter and getter methods. In that interface, core components and 
metadata are updated via replacement methods (requiring knowledge of the
class components), while our interface requires only a single call to `mutate()`
to perform the modification.

Workflows can be composed by chaining verbs together into 'sentences' via the 
forward pipe operator,`%>%`  (exported from the R package `magrittr` 
@R-magrittr), which can be read as the word 'then'. Overall, this  allows users 
to write human readable code because workflows describe what the code is doing 
rather than how its doing it.
